The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2020-07-29 Writeup/manuscript.tex
significance, abstract, intro

\title{Template for preparing your research report submission to PNAS using Overleaf}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Wallpaper groups were discovered in the mid-19th century, and the 17 groups constitute the complete set of possible ways of regularly tiling the 2D-plane. In recent years wallpaper groups have found use in the vision science community, as an ideal stimulus set for studying the perception of symmetries in textures. Here we present brain imaging and psychophysical data on the complete set of wallpaper groups and show the hierarchical organization among wallpaper groups in reflected in both visual cortex activity and performance on a symmetry detection task.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any conflict of interest here.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: author.two\@email.com}

@@

\begin{abstract}
A century of vision research has demonstrated that symmetry contributes to numerous domains of visual perception \cite{RN1311, RN1166,RN1682,RN1824}. In a 2-D image, the four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in 17 distinct ways. These 17 “wallpaper” groups \cite{RN1562,RN1563,RN1425} obey a hierarchy, determined by mathematical group theory, where simpler groups are subgroups of more complex ones\cite{RN1711}. Here we probe representations of symmetries in wallpaper groups using two methods: Steady-State Visual Evoked Potentials (SSVEPs) recorded using EEG and psychophysical detection thresholds. We find that the hierarchical relationships between the wallpaper groups are almost perfectly preserved in both behavior and response amplitudes in visual cortex. This remarkable consistency between the structure of symmetry representations and mathematical group theory, is likely generated over visual development, through implicit learning of regularities in the environment. 
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry,

In the two spatial dimensions relevant for images, the four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in 17 distinct ways, the “wallpaper” groups \cite{RN1562,RN1563,RN1425}. The 17 wallpaper groups obey a hierarchy of complexity, determined by mathematical group theory, where simpler groups are sub-groups of more complex ones \cite{RN1711}. Here we use Steady-State Visual Evoked Potentials (SSVEPs) to measure responses to 16 out of 17 wallpaper groups and show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups. Specifically, the amplitudes of symmetry-specific responses in individual participants ($n=25$) preserve these relationships at an above-chance level in 88.3\% (53 out of 60) of cases. Visual cortex thus encodes all the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory. Given that most participants had no knowledge of group theory, the ordered structure of visual responses to wallpaper groups is likely learned implicitly from regularities in the visual environment.

@@

The visual stimuli for our experiment were multiple exemplar images belonging to each of the 17 wallpaper groups, generated from random-noise textures, as described in detail elsewhere \cite{RN1725}. To isolate brain activity specific to the symmetry structure in the images from activity associated with modulation of low-level features, we used a steady-state design, in which exemplar images belonging to 16 of the 17 wallpaper groups alternated with phase-scrambled images of the same group. Because all wallpapers are periodic images due to their lattice tiling structure, the phase-scrambled images are also a wallpaper group (P1). P1 contains no symmetries other than translation, while all other groups contain translation in combination with one or more of the other three fundamental symmetries (reflection, rotation, glide reflection) \cite{RN1425}. 
Exemplars from each of the 16 groups alternated at 0.83 Hz with their corresponding set of P1 exemplars, that were matched in terms of their Fourier power spectrum. Because the P1 group serves a control stimulus in this approach, the experiment was restricted to the remaining 16 groups. This design allows us to isolate responses to structural features beyond the shared power spectrum, including any symmetries other than translation, in the odd harmonics of the image update frequency \cite{RN1725,RN1540} [NORCIA 2002 MISSING]. Thus, the magnitude of the odd harmonic response components can be used as a distance metric for each group, with distance being measured relative to the simplest group, P1. A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \cite{RN1425}. Wallpaper groups differ in the number and kind of these transformations. In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \cite{RN1425}. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the outer group \cite{RN1425}. As an example, let us consider groups P6 and P2. If we ignore the translations in two directions that both groups share, group P6 consists of the set of rotations {$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$}, in which P2 {$0^{\circ}$, $180^{\circ}$} is contained. P2 is thus a subgroup of P6, and the full P6 set can be generated by every combination of P2 and rotations {$0^{\circ}$, $120^{\circ}$, $240^{\circ}$}. Because P2 is repeated three times in P6, P2 is a subgroup of P6 with index 3 \cite{RN1425}.

@@

>>>>

\title{The human visual system preserves the hierarchy of 2-dimensional pattern regularity}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Wallpaper groups were discovered in the mid-19th century, and the 17 groups constitute the complete set of possible ways of regularly tiling the 2D-plane. In recent years wallpaper groups have found use in the vision science community, as an ideal stimulus set for studying the perception of symmetries in textures. Here we present brain imaging and psychophysical data on the complete set of wallpaper groups and show the hierarchical organization among wallpaper groups in reflected in both representations in visual cortex and performance on a symmetry detection task. This shows that the visual system is highly sensitive to reguarities in textures, and suggest that symmetries may play an important role in texture perception.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{PJK and ADFC designed the study, PJK collected EEG data, ADFC collected psychophysical data, PJK and ADFC wrote the paper.}
\authordeclaration{The authors have no conflicts of interests to declare}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: \href{mailto:pjkohler@syorku.ca}{pjkohler@yorku.ca}}

@@

\begin{abstract}
A century of vision research has demonstrated that symmetry contributes to numerous domains of visual perception \cite{RN1311, RN1166,RN1682,RN1824}. In a 2-D image, the four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in 17 distinct ways. These 17 “wallpaper” groups \cite{RN1562,RN1563,RN1425} obey a hierarchy, determined by mathematical group theory, in which simpler groups are subgroups of more complex ones\cite{RN1711}. Here we probe representations of symmetries in wallpaper groups using two methods: (1) Steady-State Visual Evoked Potentials (SSVEPs) recorded using EEG and (2) symmetry detection thresholds measured psychophysically. We find that hierarchical relationships between the wallpaper groups are almost perfectly preserved in both behavior and response amplitudes in visual cortex. This remarkable consistency between the structure of symmetry representations and mathematical group theory, is likely generated over visual development, through implicit learning of regularities in the environment. 
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four fundamental symmetries can be combined in 17 distinct ways, the “wallpaper” groups \cite{RN1562,RN1563,RN1425}. Previous work has focused on four of the wallpaper groups, and used functional MRI to show that rotation symmetries within wallpapers are represented parametrically in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect is also robust in electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend on this work by collecting SSVEPs and psychophysical data from human participants viewing the complete set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus, with the goal of providing a more complete picture of how wallpaper groups are represented in the human visual system. The 17 wallpaper groups obey a hierarchy of complexity, determined by mathematical group theory, where simpler groups are sub-groups of more complex ones \cite{RN1711}. The two datasets presented here puts on in the position of being able to assess the extent to which both behavior and brain responses follow that hierachy. The results show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds preserving these relationships at a level that is far beyond chance. Visual cortex thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory. Given that most participants had no knowledge of group theory, the ordered structure of visual responses to wallpaper groups is likely learned implicitly from regularities in the visual environment.

@@

The visual stimuli for our experiment were multiple exemplar images belonging to each of the 17 wallpaper groups, generated from random-noise textures, as described in detail elsewhere \cite{RN1725}. To isolate brain activity specific to the symmetry structure in the images from activity associated with modulation of low-level features, we used a steady-state design, in which exemplar images belonging to 16 of the 17 wallpaper groups alternated with phase-scrambled images of the same group. Because all wallpapers are periodic images due to their lattice tiling structure, the phase-scrambled images are also a wallpaper group (P1). P1 contains no symmetries other than translation, while all other groups contain translation in combination with one or more of the other three fundamental symmetries (reflection, rotation, glide reflection) \cite{RN1425}. 
Exemplars from each of the 16 groups alternated at 0.83 Hz with their corresponding set of P1 exemplars, that were matched in terms of their Fourier power spectrum. Because the P1 group serves a control stimulus in this approach, the experiment was restricted to the remaining 16 groups. This design allows us to isolate responses to structural features beyond the shared power spectrum, including any symmetries other than translation, in the odd harmonics of the image update frequency \cite{RN1725,RN1540, RN1954}. Thus, the magnitude of the odd harmonic response components can be used as a distance metric for each group, with distance being measured relative to the simplest group, P1. A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \cite{RN1425}. Wallpaper groups differ in the number and kind of these transformations. In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \cite{RN1425}. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the outer group \cite{RN1425}. As an example, let us consider groups P6 and P2. If we ignore the translations in two directions that both groups share, group P6 consists of the set of rotations {$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$}, in which P2 {$0^{\circ}$, $180^{\circ}$} is contained. P2 is thus a subgroup of P6, and the full P6 set can be generated by every combination of P2 and rotations {$0^{\circ}$, $120^{\circ}$, $240^{\circ}$}. Because P2 is repeated three times in P6, P2 is a subgroup of P6 with index 3 \cite{RN1425}.

@@

\section*{snippets}
Specifically, the amplitudes of symmetry-specific responses in individual participants ($n=25$) preserve these relationships at an above-chance level in 88.3\% (53 out of 60) of cases.
